\documentclass{tufte-handout}
\usepackage{amsmath, amsfonts, amsthm, amssymb}
\usepackage{thmtools}


%% environments
\theoremstyle{definition} \newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\theoremstyle{definition} \newtheorem{remark}{Remark}

% theorem style for example
\declaretheoremstyle[
spaceabove=6pt, spacebelow=6pt,
headfont=\normalfont\bfseries,
notefont=\mdseries, notebraces={(}{)},
bodyfont=\normalfont,
postheadspace=1em,
%numberwithin=section
]{exstyle}

% theorem style for solution
\declaretheoremstyle[
spaceabove=6pt, spacebelow=6pt,
headfont=\normalfont\bfseries,
notefont=\mdseries, notebraces={(}{)},
bodyfont=\normalfont,
postheadspace=1em,
headpunct={},
qed= $\blacksquare$,
numbered=no
]{solstyle}

\declaretheorem[style=exstyle]{example}
\declaretheorem[style=solstyle]{solution}


%% commands
\newcommand{\prob}[1]{\mathbf{P} \left( #1 \right)}
\newcommand{\cprob}[2]{\mathbf{P} \left( #1 \, | \, #2 \right)}
\newcommand{\pmf}[2]{p_{#1} \left( #2 \right)}
\newcommand{\expt}[1]{\mathbf{E} \left[ #1 \right]}
\newcommand{\jpmf}[4]{p_{#1, #2} \left( #3, #4 \right)}
\newcommand{\cpmf}[3]{p_{#1 | #2} \left( #3 \right)}
\newcommand{\cexpt}[2]{\mathbf{E} \left[ #1 \, | \, #2 \right]}


\author{John N. Tsitsiklis}
\title{Counting}
\begin{document}
\maketitle
\section{Basics}

\subsection{The Counting Principle}
Consider a process that consists of $r$ stages. Suppose that:
\begin{enumerate}[(a)]
\item There are $n_1$ possible results at the first stage.
\item For every possible result at the first stage, there are $n_2$
  possible results at the second stage.

\item More generally, for an sequence of possible results at the first
  $i - 1$ stages, there are $n_i$ possible results at the $i^{\text{th}}$
  stage. Then, the total number of possible results of the $r$-stage
  process is
  \begin{equation*}
    n_1 n_2 \cdots n_r
  \end{equation*}
\end{enumerate}

\subsection{$k$-permutations}
We start with $n$ distinct objects, and let $k$ be some positive integer,
with $0 \leq k \leq n$. We wish to count the number of different ways
that we can pick $k$ out of these $n$ objects and arrange them in a
sequence.
\begin{equation*}
  k\text{-permutations} = \frac{n!}{(n-k)!}
\end{equation*}

\subsection{Combinations}
We start with $n$ distinct objects, and let $k$ be some positive integer,
with $0 \leq k \leq n$. We wish to count the number of $k$-element
subsets of a given $n$-element set. The number of possible combinations
is
\begin{equation*}
  \frac{n!}{k!(n-k)!} \text{ which is same as } \binom{n}{k}
\end{equation*}

\subsection{Partitions}
We are given an $n$-element set and non-negative integers
$n_1, n_2, \ldots, n_r$, whose sum is equal to $n$. We consider partitions
of the set into $r$ disjoint subsets, with the $i^{\text{th}}$ subset
containing exactly $n_i$ elements. The number of partitions is
\begin{equation*}
  \frac{n!}{n_1! n_2! \cdots n_r!} \text{ which is denoted by }
  \binom{n}{n_1, n_2, \ldots, n_r}
\end{equation*}

\subsection{Integer Partitions}
Given a non-negative integer $n$ and a positive integer $k$, consider the
equation
\begin{equation*}
  x_1 + x_2 + \cdots + x_k = n
\end{equation*}
to be solved with respect to non-negative integer variables
$x_1, x_2, \ldots, x_k$. Find the total number of solutions?

There is a one-to-one correspondence between non-negative integer
solutions of equation $x_1 + x_2 + \cdots + x_k = n$ and sequences of
$n + k - 1$ symbols ($n$ ''$o$'' and $k - 1$ ''$|$''). Hence the total
number of solutions equals the number of ways of selecting $k - 1$ places
for the symbol ''$|$'' in a sequence of length $n + k - 1$. Hence the
total number of solutions
\begin{equation*}
  \binom{n + k - 1}{k - 1}
\end{equation*}
\section{Examples}
\begin{example}
  Ninety students, including Joe and Jane, are to be split into three
  classes of equal size, and this is to be done at random. What is the
  probability that Joe and Jane end up in the same class?
\end{example}

\begin{solution}
  We place Joe in one class. Regarding Jane, there are $89$ possible
  slots, and only $29$ place her in the same class as Joe. Thus the
  answer is $\frac{29}{89}$.
\end{solution}

 \begin{example}[Hypergeometric probabilities]
   An urn contains $n$ balls, out of which $m$ are red. We select $k$ of
   the balls at random, without replacement. What is the probability that
   $i$ of the selected balls are red?
 \end{example}

 \begin{solution}
   The sample space consists of $\binom{n}{k}$ different ways of
   selecting $k$ out of $n$ balls. For the event of interest to occur,
   we have to select $i$ out of the $m$ red balls and also select $k - i$
   balls out of the $n - m$ balls, which are not red. Thus the desired
   probability is
   \begin{equation*}
     \frac{\binom{m}{i} \binom{n - m}{k - i}}{\binom{n}{k}}
   \end{equation*}
   For $i \geq 0$ satisfying $i \leq m, i \leq k$ and $k - i \leq n - m$.
   For all other $i$, the probability is zero.
 \end{solution}

 \begin{example}
   Prove the algebraic identity
   \begin{equation*}
     \sum_{k = 1} ^n k \binom{n}{k} = n 2^{n - 1}
   \end{equation*}
   
 \end{example}
 \begin{solution}
   We have a group of $n$ persons. Consider clubs that consists of a
   special person from the group (the club leader) and a number (possibly
   zero) of additional club members. Let us count the number of possible
   clubs of this type in two different ways.

   There are $n$ choices for club leader. Once the leader is chosen, we
   are left with a set of $n - 1$ available persons, and we are free to
   choose any of the $2^{n - 1}$ subsets. Thus the number of possible
   clubs is $n 2^{n - 1}$.

   Alternatively, for fixed $k$, we can form a $k$-person club by first
   selecting $k$ out of the $n$ available persons, which can be done
   in $\binom{n}{k}$ ways. We can then select one of the members to
   be the leader, for which we have $k$ choices. Hence the total mumber
   of clubs is $\sum_{k = 1} ^n k \binom{n}{k}$
   \begin{equation*}
     \sum_{k = 1} ^n k \binom{n}{k} = n 2^{n - 1}
   \end{equation*}
 \end{solution}
\end{document}
